---
title: 'Lab 5 Part 2: Spatial data wrangling, visualization, and a variogram'
author: "Kirsten Hodgson"
date: "2/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(fable)
library(sf)
library(tmap)
```

### California county outlines (polygons)

```{r}
ca_counties <- read_sf(here("data", "ca_counties", "CA_Counties_TIGER2016.shp"))

ca_subset <- ca_counties %>% 
  select(NAME, ALAND) %>% 
  rename(county_name = NAME, land_area = ALAND)
```

A sf (simple features) object assumes you want to keep the spatial information, so the geometry column is retained even though I didn't select for it during wrangling (called "sticky geometry")

#### Check and set CRS

- Using `st_crs()` to check the existing CRS for spatial data. We see that this CRS is WGS84 (epsg: 3857)

```{r}
ca_subset %>% st_crs()
```

#### Look at it

Plot the California counties using  `geom_sf()`. Notice that we can update aesthetics just like we would for a regular ggplot object. Here, we update the color based on land area (and change the color gradient).

```{r}
ggplot(data = ca_subset) +
  geom_sf(aes(fill = land_area), color = "white", size = 0.1) +
  theme_void() +
scale_fill_gradientn(colors = c("cyan", "blue", "purple"))
```

### Invasive red sesbania records (spatial points)

Red sesbania (*Sesbania punicea*) is an invasive plant. Observations for locations of invasive red sesbania are from CA DFW.

The data exist `data/red_sesbania`, and the shapefile is stored as `ds80.shp`. Let's read in the data:

```{r}
sesbania <- read_sf(here("data", "red_sesbania", "ds80.shp"))

#Check the CRS:
sesbania %>% st_crs()
```

This CRS is different from the California counties CRS, so we need to update it to match. Use `st_transform()` to update the CRS:

```{r}
sesbania <- st_transform(sesbania, 3857)

#Then check it:
sesbania %>% st_crs()
```

#### Plot them together!

```{r}
ggplot() +
  geom_sf(data = ca_subset) +
  geom_sf(data = sesbania, size = 1, color = "red")
```

#### A bit of wrangling!

If we want to find the count of red sesbania observed locations in this dataset *by county*: Use `st_join()` to do a spatial join!

```{r}
ca_sesbania <- ca_subset %>% 
  st_join(sesbania)
```

And then we can find counts (note: these are not counts for individual plants, but by record in the dataset) by county:

```{r}
sesbania_counts <- ca_sesbania %>% 
  count(county_name)
```

Then we can plot a chloropleth using the number of records for red sesbania as the fill color (instead of what we used previously, land area):

```{r}
ggplot(data = sesbania_counts) +
  geom_sf(aes(fill = n), color = "white", size = 0.1) +
  scale_fill_gradientn(colors = c("lightgray", "orange", "red")) +
  theme_minimal() +
  labs(fill = "Number of S. punicea records")
```

So we see that we can still use our usual wrangling skills! Let's do a bit more for fun, just to prove that our existing wrangling skills still work with spatial data - the spatial information just sticks to it! Only plot the county with the greatest number of red sesbania records (Solano), and make a map of those locations (yeah there are many ways to do this):

```{r}
#Subset of sesbania point locations only in Solano County
solano_sesbania <- sesbania %>% 
  filter(COUNTY == "Solano")

#Only keep Solano polygon from California County data
solano <- ca_subset %>% 
  filter(county_name == "Solano")

ggplot() +
  geom_sf(data = solano) +
  geom_sf(data = solano_sesbania)
```

### Making an interactive map with {tmap}

Sometimes we'll want to make a map interactive so that audience members can zoom in, explore different areas, etc. We can use the {tmap} package to create an interactive map. Let's make one for our California counties (fill aesthetic by land area) with the red sesbania locations on top:

```{r}
#Set the viewing mode to "interactive":
tmap_mode(mode = "view")

#Then make a map (with the polygon fill color updated by variable `land_area`, updating the color palette to "BuGn"), then add another shape layer for the sesbania records (added as dots):
tm_shape(ca_subset) +
  tm_fill("land_area", palette = "BuGn") +
  tm_shape(sesbania) +
  tm_dots()
```

See *all* kinds of other cool ways you can update your interactive tmaps.

See:

- `tmap` vignettes
- Chapter 8 in Robin Lovelace's "Geocomputation in R"